Behind Bars and Beyond: Investigating Socioeconomic Drivers of Incarceration Rates in the United States

### 2024-04-18 paper/paper.qmd
Finalized paper and all references.
Completed the discussion and limitations sections. Added an appendix with model checks, and added all references as well as the abstract.

---
title: "Behind Bars and Beyond: Investigating Socioeconomic Drivers of Incarceration Rates in the United States"
subtitle: "A comparative analysis of global trends and America's mass incarceration crisis."
author: 
  - Abbass Sleiman
thanks: "Code and data are available at: [https://github.com/AbbassSleiman/US-Middle-Class](https://github.com/AbbassSleiman/US-Middle-Class)."
date: today
date-format: long
format: 
  pdf: 
    include-in-header: 
      text: |
        \usepackage{lscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
    documentclass: article
    geometry: margin = 1.2in
    abstract: "Over the past five decades, the United States has witnessed a decline in the percentage of adults residing in middle-class households, alongside escalating concerns about housing affordability. This study aims to assess the impact of rising costs on the living standards of middle and lower-income households, recognizing them as the most affected demographic by these economic shifts. Utilizing data from [...], we analyze changes in income distribution and evaluate the influence of economic policy changes, including interest rates and taxation, on the standards of living of these individuals. Additionally, we explore how individual-level characteristics intersect with these policies to address how best to pursue the goal of economic equity across individuals of all backgrounds. We find that [...]."
    number-sections: true
output: 
  bookdown::pdf_document2
toc: true
bibliography: references.bib
---

@@

# Package loading
library(tidyverse)
library(here)
library(ggplot2)
library(knitr)
library(modelsummary)
library(rstanarm)
library(kableExtra)

@@

This paper utilizes various datasets from the World Population Review, in particular it utilizes their datasets on "Incarceration Rates by Country 2024", "Poverty Rate by Country 2024", "Education Rankings by Country 2024", "Unemployment by Country 2024", and "Violent Crime Rates by Country 2024". By conducting statistical analyses with the aforementioned data, this paper finds evidence to suggest that [...]

@@

All data analysis was done through R [@citeR] with the aid of the following packages:

@@

## Model Coefficients

@@

## Actual vs. Predicted Incarceration Rates

@@

## Nations with extreme incarceration rates necessitate the need to consider additional factors beyond the model's scope

As the results in @sec-results have shown, nations with excessively high and excessively low incarceration rates have various qualities about them, outside the scope of our model, that may better explain how and why they may find themselves with the incarceration rates that they do have. In particular, various cultural differences across nations such as differences in how nations view the idea of incarceration could significantly impact how various countries choose to deal with crime.

@@

## Financial hardships and corruption at the national level may prevent nations from incarcerating when necessary

@@

## Third discussion point

# Weaknesses and next steps {#sec-limitations}

(Limitation, could be better to group nations into groups where they share similar cultural values and whatnot in order to account for differences in culture.)

@@

# Additional data details

# Model details {#sec-model-details}

@@

```{r}
#| label: fig-ppcheck-postprior
#| echo: false
#| fig-pos: H
#| message: false

pp_check(analysis_model) + theme_classic()

posterior_vs_prior(analysis_model) + theme_classic() + coord_flip()
```

```{r}
#| label: fig-rhat
#| echo: false
#| fig-pos: H

plot(analysis_model, "rhat")
```
\newpage

>>>>

---
title: "Behind Bars and Beyond: Investigating Socioeconomic Drivers of Incarceration Rates in the United States"
subtitle: "A comparative analysis of global trends and America's mass incarceration crisis."
author: 
  - Abbass Sleiman
thanks: "Code and data are available at: [https://github.com/AbbassSleiman/US-Middle-Class](https://github.com/AbbassSleiman/US-Middle-Class)."
date: today
date-format: long
format: 
  pdf: 
    include-in-header: 
      text: |
        \usepackage{lscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
    documentclass: article
    geometry: margin = 1.2in
    abstract: "Over the last 2 decades, the United States has found itself struggling with an issue of mass incarceration, leading the world in number of individuals incarcerated. This paper investigates the possible causes of this by conducting a comparative analysis of global trends in education, unemployment, crime, and poverty rates. Utilizing data from the World Population Review, and employing a Bayesian multiple linear regression model, this paper finds evidence that poverty and unemployment negatively correlate with incarceration rates, whereas violent crime rates positively correlate. The paper also identifies nuances in the relationship between educational attainment and incarceration rates. These insights may offer valuable considerations for policy-making aimed at addressing mass incarceration in the United States."
    number-sections: true
output: 
  bookdown::pdf_document2
toc: true
bibliography: references.bib
---

@@

# Package loading
library(tidyverse)
library(here)
library(ggplot2)
library(knitr)
library(modelsummary)
library(rstanarm)
library(kableExtra)
library(dplyr)
library(janitor)

@@

This paper utilizes various datasets from the World Population Review, in particular it utilizes their datasets on "Incarceration Rates by Country 2024", "Poverty Rate by Country 2024", "Education Rankings by Country 2024", "Unemployment by Country 2024", and "Violent Crime Rates by Country 2024". By conducting statistical analyses with the aforementioned data, this paper finds evidence to suggest that poverty and unemployment negatively correlate with incarceration rates, and violent crime rates correlate positively. Education, on the other hand, appears to have a more nuanced relationship with incarceration rates as the relationship does not appear to be consistent across all levels of education.

@@

All data analysis was done through R [@citeR] with the aid of the following packages: `tidyverse` [@citetidyverse], `here` [@citehere], `ggplot2` [@citeggplot2], `knitr` [@citeknitr], `modelsummary` [@citemodelsummary], `rstanarm` [@citerstanarm], `kableExtra` [@citekableExtra], `dplyr` [@citedplyr], and `janitor` [@citejanitor].

@@

## Model Coefficients {#sec-modelcoefficients}

@@

## Actual vs. Predicted Incarceration Rates {#sec-actualvspredicted}

@@

## Nations with extreme incarceration rates necessitate the need to consider additional factors beyond the model's scope {#sec-culture}

As the results in @sec-actualvspredicted have shown, nations with excessively high and excessively low incarceration rates have various qualities about them, outside the scope of our model, that may better explain how and why they may find themselves with the incarceration rates that they do have. In particular, various cultural differences across nations such as differences in how nations view the idea of incarceration could significantly impact how various countries choose to deal with crime.

@@

## Economic hardships and corruption at the national level may prevent nations from incarcerating when necessary {#sec-povertycorruption}

@@

## Educational attainment as a nuanced metric of incarceration rates {#sec-educationincarceration}

As was explored in @sec-modelcoefficients, we found that there was no consistent relationship between educational ranking and incarceration rates. In particular, though the highest educationally ranked nations found themselves with lower incarceration rates than those with poorer education, the extent to which incarceration rates were predicted to rise with a decrease in education varied depending on how much the educational ranking would fall. Somewhat counter intuitively, a nation falling from the highest educational category to that of the ranks "111-125" would find themselves with the highest increase in incarceration rates compared to a further worsening of education to even the lowest category of "176-198".

The possible explanations for this are twofold: Firstly, as was previously examined in @sec-culture, incarceration rates can be heavily dependent on cultural aspects that were not taken into account via the model. As such, certain nations with exceedingly low or exceedingly high incarceration rates may be disproportionately affecting the mean incarceration rates of their respective educational ranking categories, in turn leading to inaccurate conclusions regarding the impact of education on incarceration rates. To illustrate this point further, El Salvador who feature the highest incarceration rate of 1086, also happen to be in the educational ranking category of "111-125", which may explain why the model predicts that this particular category is likely to raise a nation's incarceration rate by the greatest margin.

The second possible explanation could be that nations with poorer levels of education are likely to also be poorer economically, which, as explored in @sec-povertycorruption, could lead to lower incarceration rates, in turn providing a possible explanation for why nations with the lowest levels of education also happen to feature lower incarceration rates than some nations with better education.

## Insights for addressing the mass incarceration challenge in the United States drawn from the data and results

By incorporating all insights derived from @sec-actualvspredicted, @sec-povertycorruption, and @sec-educationincarceration, we arrive to the following conclusions for how the United States can best address their issue of mass incarceration:

(1) The United States should stray from punitive measures as the primary means of dealing with crime as this can perpetuate a cycle of high incarceration rates, recidivism, and crime. Instead, shifting towards rehabilitation-oriented policies, such as those endorsed by nations like Norway, can be more effective in reducing the need for incarceration, and reduce the likelihood of repeat-offenders.

(2) The United States should dedicate more resources towards enhancing their educational prowess. As it stands, the United States has an education ranked between "20-34", and as a result, given the overwhelming evidence that nations with the highest educational strength also happen to have the lowest incarceration rates, an increase in investment in education can aid the United States in decreasing both their violent crime rate, as well as the need for mass incarceration. Though we have uncovered that education can be closely linked to other factors of a nation including economic strength and standards of living, the United States is one of the most economically powerful nations, and as a result, we may come to the justified conclusion that improvements in their education are likely to yield positive results akin to that of similar nations.

# Limitations and Next Steps {#sec-limitations}

## Limitations

There are a number of limitations to be addressed with regards to the analysis conducted and subsequent conclusions drawn. A significant limitation pertaining to this study is the lack of up-to-date data across all nations. As was discuss is @sec-data, all raw datasets contained information on nations that may not be accurate as of 2024. In particular, many nations had data that was collected years, and in some cases decades, ago, in turn potentially rendering much of the analysis inaccurate as significant changes could have risen over the years. Moreover, not all datasets included information on all nations, and in many cases, countries that were included in some datasets featured missing data points for various qualities of differing nations. As a result, the subset of countries that were included in this study may not accurately reflect the overarching global trends as only 134 out of the 193 countries recognized by the UN [@citeUN] were taken into account during the analysis. Thus, the subsequent conclusions drawn may be biased as the remaining countries left in the study could be skewed towards more developed nations as these may be more likely to have data readily available. Moreover, due to the missing data points for countries that were included in the study, values such as the violent crime rate statistic that was calculated for each nation is inaccurate as it only included whatever information was available on that particular nation. As a result, nations which may have reported more crime rates than others would in turn bias the results to inaccurately portray such nations as having relatively more crime than they should when compared with nations that had missing data points for certain crime statistics.

A potentially significant limitation also exists with regards to the model used. In particular, the model does not take into account any interaction terms between the various independent variables when it likely should have, given the complex and intricate dynamics between the many variables at play. Though this was done to simplify the complexity of the model to allow for easier interpretations of the results, the caveat was an overall less realistic, and possibly less accurate, model. Another limitation with the model lies in its omission of cultural variations and associated factors across nations. By not grouping countries with similar cultural values or political and judicial policies, the model fails to account for factors that could greatly influence the derived results. As we have already explored in @sec-culture, varying cultural and societal norms with regards to how nations deal with crime can significantly impact their incarceration rates, a fact not taken into account by the model.

## Next Steps

Future analysis in the context of this study can be better improved by incorporating more up-to-date data. In particular, web-scraping and compiling the raw data manually as opposed to using the raw datasets used in this study could net an overall more recent, and complete, set of data points. Moreover, incorporating time series data spanning multiple years across all nations would allow us to account for time and country fixed-effects, in turn providing more conclusive data that takes into account the various cultural and societal disparities across nations, and years. Another improvement that could be made is a better analysis of the model itself in order to fine-tune aspects such as the locations of the various priors, as well as exploring the possibility of adding additional factors and interaction terms to more accurately model incarceration rates.

@@

# Model details

@@

We employ the use a posterior predictive check in @fig-ppcheck to evaluate how well the model fits the data. By examining the figure, it appears as though the model does not accurately fit the data, and thus amendments likely should be made as a result [@citeAlexander].

```{r}
#| label: fig-ppcheck
#| fig-cap: Examining how the model fits, and is affected by, the data through a posterior predictive check
#| echo: false
#| fig-pos: H
#| message: false
#| warning: false

pp_check(analysis_model) + 
  theme_classic() +
  theme(legend.position = "bottom")
```

## Comparison of the posterior and prior

We compare the posterior and prior in @fig-postprior to examine how the estimates change once data is taken into account [@citeAlexander].

```{r}
#| label: fig-postprior
#| fig-cap: Examining how the model fits, and is affected by, the data through a posterior-prior comparison
#| echo: false
#| warning: false
#| fig-pos: H
#| message: false

posterior_vs_prior(analysis_model) + 
  theme_classic() + 
  theme(legend.position = "bottom") +
  coord_flip()
```

## Markov Chain Monte Carlo Convergence

As `rstanarm` makes use of the sampling algorithm MCMC [@citeAlexander], we examine whether the algorithm ran into issues through the use of an Rhat plot and trace plot, both showcased in @fig-mcmc. The Rhat plot does not indicate any issues, evidenced by the values all remaining close to 1 [@citeAlexander]. Moreover, the trace plot suggests nothing out of the ordinary given that lines remain horizontal and fluctuate in an expected manner [@citeAlexander].

```{r}
#| label: fig-mcmc
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Rhat plot", "Trace plot"]
#| echo: false
#| message: false
#| warning: false
#| fig-pos: H
#| layout-ncol: 1

plot(analysis_model, "rhat") + 
  theme_classic()

plot(analysis_model, "trace")

## Credibility Intervals

Lastly, we employ a 95% credibility interval as showcased in @fig-ci, to gain a better understanding of the probability distribution of our coefficients.

```{r}
#| label: fig-ci
#| fig-cap: "Credible intervals"
#| echo: false
#| message: false
#| warning: false
#| fig-pos: H

plot(analysis_model) + 
  theme_classic()
